The Leaky Semicolon

### 2021-10-29 paper/intro.tex
pass over intro

In this paper, we show that \PwP{} can be extended with \emph{families of
  predicate transformers} (\PwT{}) to calculate sequential dependencies in a way that
is \emph{compositional} and \emph{direct}: \emph{compositional} in that the
denotation of $(\aCmd_1\SEMI \aCmd_2)$ can be computed from the denotation of
$\aCmd_1$ and the denotation of $\aCmd_2$, and \emph{direct} in that these
can be calculated independently.  With this formulation, we can show:
\begin{align*}
  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp} &=
  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp} \;\;\text{if}\; \aLoc\neq\bLoc
\end{align*}
Then the equivalence holds in any context---this form of the equivalence
enables reasoning about peephole optimizations.  Unlike prior work, \PwT{}
allows the presence or absence of a dependency can be understood in
isolation.  This enables incremental and modular validation of assumptions
about program dependencies in larger blocks of code.

Our main insight is that for language models, \emph{sequentiality} is the
hard part.  \emph{Concurrency} is easy!  Or at least, it is no more difficult
than it is for hardware.  Compilers make the difference, since they typically
do little optimization between threads.  We motivate our approach to
sequential dependencies in \textsection\ref{sec:overview} and provide formal
definitions in \textsection\ref{sec:model}.  We extend the model to include
additional features, such as address calculation and \RMW{}s, in
\textsection\ref{sec:additional}.  We further discuss related work in
\textsection\ref{sec:related}.

We extend \PwT{} to a full memory model in \textsection\ref{sec:mca}, based
on \PwP{} \cite{DBLP:journals/pacmpl/JagadeesanJR20}.
\textsection\ref{sec:results} summarizes the results for this model.  The
dependency relation calculated by \PwT{} can also be used with off-the-shelf
models.  For example, in \textsection\ref{sec:c11} we show that it can be
used as an $\rsdep$ relation for \cXI, adapting the approach of \MRD{}
\cite{DBLP:conf/esop/PaviottiCPWOB20}.  % This is only the second semantics
% to distill dependencies down to a relation compatible with the existing
% \cpp{} standard, and the first one to be fully compositional.
\textsection\ref{sec:tool} describes a tool for automatic evaluation of
litmus tests in this model.  \cXI{} allows thin-air in order to avoid
overhead in the implementation of relaxed reads.  Safe language like OCaml
\cite{Dolan:2018:BDR:3192366.3192421} have typically made the opposite choice.  Just
\PwT{} can be used to strengthen \cXI{}, it could also be used to weaken these
models, allowing optimal lowering for relaxed reads while banning thin-air.

@@

Supplementary material for this paper available at
\url{https://weakmemory.github.io/pwt}.

>>>>

In this paper, we show that \PwP{} can be extended with \emph{families of
  predicate transformers} (\PwT{}) to calculate sequential dependencies in a way that
is \emph{compositional} and \emph{direct}: \emph{compositional} in that the
denotation of $(\aCmd_1\SEMI \aCmd_2)$ can be computed from the denotation of
$\aCmd_1$ and the denotation of $\aCmd_2$, and \emph{direct} in that these
can be calculated independently.  With this formulation, we can show:
\begin{align*}
  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp} &=
  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp} \;\;\text{if}\; \aLoc\neq\bLoc
\end{align*}
Then the equivalence holds in any context---this form of the equivalence
enables reasoning about peephole optimizations.  Said differently, unlike prior work, \PwT{}
allows the presence or absence of a dependency to be understood in
isolation---this enables incremental and modular validation of assumptions
about program dependencies in larger blocks of code.

Our main insight is that for language models, \emph{sequentiality} is the
hard part.  \emph{Concurrency} is easy!  Or at least, it is no more difficult
than it is for hardware.  Compilers make the difference, since they typically
do little optimization between threads.  We motivate our approach to
sequential dependencies in \textsection\ref{sec:overview} and provide formal
definitions in \textsection\ref{sec:model}.  We extend the model to include
additional features, such as address calculation and \RMW{}s, in
\textsection\ref{sec:additional}.  We discuss related and future work in
\textsection\ref{sec:related}--\ref{sec:outro}.

We extend \PwT{} to a full memory model in \textsection\ref{sec:mca}, based
on \PwP{} \cite{DBLP:journals/pacmpl/JagadeesanJR20}.
\textsection\ref{sec:results} summarizes the results for this model.  In
addition to powering such a bespoke model, the dependency relation calculated
by \PwT{} can also be used with off-the-shelf models.  For example, in
\textsection\ref{sec:c11} we show that it can be used as an $\rsdep$ relation
for \cXI, adapting the approach of \MRD{}
\cite{DBLP:conf/esop/PaviottiCPWOB20}.  % This is only the second semantics
% to distill dependencies down to a relation compatible with the existing
% \cpp{} standard, and the first one to be fully compositional.
\textsection\ref{sec:tool} describes a tool for automatic evaluation of
litmus tests in this model.  \cXI{} allows thin-air in order to avoid
overhead in the implementation of relaxed reads.  Safe language like OCaml
\cite{Dolan:2018:BDR:3192366.3192421} have typically made the opposite
choice, accepting a performance penalty in order to avoid thin-air.  Just as
\PwT{} can be used to strengthen \cXI{}, it could also be used to weaken
these models, allowing optimal lowering for relaxed reads while banning
thin-air.

@@

Supplementary material for this paper is available at
\url{https://weakmemory.github.io/pwt}.
